---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
*Setting the Working Directory

```{r}
setwd("C:/Users/Melanie/Desktop/R Code Class/Reproducible Research 1/RepData_PeerAssessment1")
```

*Reading in the Data Set
```{r}
Data <- read.csv(file="C:\Users\Melanie\Desktop\R Code Class\Reproducible Research 1\RepData_PeerAssessment1\activity.zip\activity.csv"
```

*Converting the Date Field
```{r}
Data$date <- as.Date(Data$date)
```

*Load reshape2 Library
```{r}
library(reshape2)
```

*Prepare Data for use in the rest of the Assignment
```{r}
MeltDate <- melt(Data, id.vars="date", measure.vars="steps", na.rm=FALSE)
CastDate <- dcast(MeltDate, date ~ variable, sum)
```

## What is mean total number of steps taken per day?

*Show the Total # of Steps Taken Each Day
```{r}
CastDate$steps
```

*Hitogram of the # of steps taken each day
```{r}
plot(CastDate$date, CastDate$steps, type="h", main="Histogram of Daily Steps", xlab="Date", ylab="Steps per Day", col="blue", lwd=8)
abline(h=mean(CastDate$steps, na.rm=TRUE), col="red", lwd=2)
```

*Calculate the Mean of the steps taken each day
```{r}
paste("The Mean of the total # of steps taken each day is:", Mean(CastDate$steps, na.rm=TRUE))
```

*Calculate the Median of the steps taken each day

```{r}
paste("The Median of the total # of steps taken each day is:", Median(CastDate$steps, na.rm=TRUE))")
```


## What is the average daily activity pattern?
* Prep the Data to make the plot
```{r}
MeltInt <- melt(Data, id.vars="interval", measure.vars="steps", na.rm=TRUE)
CastInt <- dcast(MeltInt, interval ~ variable, mean)
```

*Create a Plot @ 5 minute intervals
```{r}
plot(CastInt$interval, CastInt$steps, type="l", main="Frequency of Steps Taken at Each Interval", xlab="Interval ID", ylab="Steps", col="orange", lwd=2)
abline(h=mean(CastInt$steps, na.rm=TRUE), col="red", lwd=2)
```

*Identify the intervial with the Max Value
```{r}
paste("Interval with max value =", CastInt$interval[which(CastInt$steps == max(CastInt$steps))])
```


## Imputing missing values
* # of Na's Calculated
```{r}
paste("The number of NAs in the data set is", sum(is.na(Data$steps)))
```

*Replace the Nas with the integer mean(CastInt Value we created earlier). To do this I created a data set containing the integer means and then using a index of NA values, I reassigned each NA values with the corresponding Integer Mean Value.
```{r}
AssignIntData = merge(Data, CastInt, by="interval", suffixes=c(".act", ".spi"))
NADataIndex = which(is.na(Data$steps))
Data[NADataIndex,"steps"] = AssignIntData[NADataIndex,"steps.spi"]
```

*redo earlier data minipulation to print a new histogram
```{r}
MeltDateNoNA <- melt(Data, id.vars="date", measure.vars="steps", na.rm=FALSE)
CastDateNoNA <- dcast(MeltDateNoNA, date ~ variable, sum)
```

*reprint Histogram
```{r}
plot(CastDateNoNA$date, CastDateNoNA$steps, type="h", main="Histogram of Daily Steps (Updated NA Values)", xlab="Date", ylab="Steps", col="gray", lwd=8)
abline(h=mean(CastDateNoNA$steps), col="red", lwd=2)
```

*Re-Calculate the Mean of the steps taken each day
```{r}
paste("The updated Mean of the total # of steps taken each day is:", Mean(CastDate$steps, na.rm=TRUE))
```

*Re-Calculate the Median of the steps taken each day

```{r}
paste("The upated Median of the total # of steps taken each day is:", Median(CastDate$steps, na.rm=TRUE))")
```

*Original Data Set (NA values left as is) - Mean daily steps = 10,766.19 - Median daily steps = 10,765

*New Data Sets (NAs imputed with mean value for that interval) - Mean daily steps = 10,890 - Median daily steps = 11,015

## Are there differences in activity patterns between weekdays and weekends?

*Tag Data as weekday or weekend
```{r}
for (i in 1:nrow(Data)) {
    if (weekdays(Data$date[i]) == "Saturday" | weekdays(Data$date[i]) == "Sunday") {
        Data$dayOfWeek[i] = "weekend"
    } else {
        Data$dayOfWeek[i] = "weekday"
    }
}
```

*subset the Data
```{r}
DataWeekday <- subset(Data, dayOfWeek=="weekday")
DataWeekend <- subset(Data, dayOfWeek=="weekend")
```

*Prep Data for Plotting
```{r}
MeltWeekday <- melt(DataWeekday, id.vars="interval", measure.vars="steps")
MeltWeekend <- melt(DataWeekend, id.vars="interval", measure.vars="steps")
CastWeekday <- dcast(MeltWeekday, interval ~ variable, mean)
CastWeekend <- dcast(MeltWeekend, interval ~ variable, mean)
```

*Load needed Plot Libraries
```{r}
library(ggplot2)
library(gridExtra)
```

*Create the plots
```{r}
WeekdayPlot <- qplot(CastWeekday$interval, CastWeekday$steps, geom="line", data=CastWeekday, type="bar", main="Steps by Interval - Weekday", xlab="Interval ID", ylab="Number of Steps")
WeekendPlot <- qplot(CastWeekend$interval, CastWeekend$steps, geom="line", data=CastWeekend, type="bar", main="Steps by Interval - Weekend", xlab="Interval ID", ylab="Number of Steps")
```
*Print the plots
```{r}
grid.arrange(WeekdayPlot, WeekendPlot, nrow=2)
```
